--- 
title: "Marine Genomics"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: ""
---

# Introduction to Marine Genomics

Rachael Bay (rbay@ucdavis.edu) 

Serena Caplins (sacaplins@ucdavis.edu)

Vanessa Guerra (vguerra4@ucmerced.edu) 

TA: Madison Armstrong (mlarmstrong@ucdavis.edu) 

 

Lecture Tuesdays 12:10-1:30, Coding session Thursdays 12:10-1:00

Office hours: Serena Caplins Thursday 1:10-2:00 PM

Madison Armstrong TBD

The **bookdown** package can be installed from CRAN or Github:

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.org/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
